---
title: "Replication Simulation"
output:
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
editor_options: 
  chunk_output_type: console
---

```{r include = FALSE}
# libraries and other things
library(PRP)
library(rstudioapi)
library(mamba)
library(data.table)
library(parallel)
library(knitr)
library(pander)

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

set.seed(2021)
```

# Common Ground

We start by generating data using the MAMBA package.
```{r}
# simulate data with MAMBA package
mamba_data <- generate_data_mamba()

# effect sizes and variance
snpeffect <- mamba_data$betajk
snpvar <- mamba_data$sjk2
```

Then we fit the simulated data to the MAMBA model. 
```{r}
# fitting simulated MAMBA data to model
if (file.exists("../data/sim_mamba_mod.rda")) {
  
  load(file = "../data/sim_mamba_mod.rda")
  
} else {
  
  sim_mod<-mamba(betajk = snpeffect, sjk2 = snpvar)
  save(sim_mod, file = "../data/sim_mamba_mod.rda")
  
}

# take values from model
pprs <- sim_mod$ppr
effectsize <- sim_mod$mu.hat
outliers <- sim_mod$outliermat[ppr < 0.5]
```

For reference, a table of the first ten SNPS and the probabilities of them being replicable is given below.
```{r echo = FALSE}
pander(outliers[1:10, ])
```

The PRP package only looks at one SNP at a time, so for comparison we'll just look at only the first SNP  The posterior-probability-of-replicability (PPR) for the first SNP is `r pprs[1]`.

```{r}
# fitting with prp library
post_prp <- posterior_prp(beta = snpeffect[1, ],
              se = sqrt(snpvar[1, ]))

```

Our posterior predictive replication p-values (posterior PRP) for the first SNP using the PRP package is `r post_prp$pvalue`.


\newpage
# Quantitative/Qualitative Differences

Now we look at a few examples.

## Case 1: Non-Zero Effect Sizes
In the first case, we use a fake model with non-zero effect sizes containing an extreme outliers.

```{r}
# high value nonzero effect size
# fake effect size
fakesnp1_1 <- c(6, 7, 6, 20)
fakesnp1_2 <- c(7, 8, 8, 9)
fakesnp1_3 <- c(7, 6, 7, 8)
fakesnpeffect1 <- rbind(fakesnp1_1, fakesnp1_2, fakesnp1_3)


# fake variance
fakesnpv1_1 <- c(3, 3, 3, 3)
fakesnpv1_2 <- c(4, 5, 6, 3)
fakesnpv1_3 <- c(4, 5, 4, 4)
fakesnpvar1 <- rbind(fakesnpv1_1, fakesnpv1_2, fakesnpv1_3)

```


```{r results = "hide"}
fakemod1 <-mamba(betajk = fakesnpeffect1,
                 sjk2 = fakesnpvar1)
```

The outlier probabilities from MAMBA are given in the table below.
```{r echo = FALSE}
pander(fakemod1$outliermat)
```

First, with the MAMBA model, we can see that the model detected an outlier in the first SNP, given by the PPR value of `r fakemod1$ppr[1]`. 

```{r}
fakemod1_post_prp <- posterior_prp(beta = fakesnpeffect1[1, ],
                                   se = sqrt(fakesnpvar1[1, ]))
```

Using the PRP package, we get a PRP value of `r fakemod1_post_prp$pvalue`.

\newpage
## Case 2: Normal Distribution, Non-Zero Effect Size SNP
In the second case, we generate a normal distribution with mean = 0 and standard deviation = 1, with one SNP having a non-zero effect size

```{r}
# generating normal data
fakesnp_norm2_1 <- rnorm(10, mean= 10, sd = 2)
fakesnp_norm2_2 <- rnorm(10, mean= 0, sd = 2)
fakesnp_norm2_3 <- rnorm(10, mean = 0, sd = 1)
fakesnpeffect2 <- rbind(fakesnp_norm2_1,
                        fakesnp_norm2_2,
                        fakesnp_norm2_3)

fakesnp_norm_var2_1 <- rnorm(10, mean = 2, sd = 0.1)
fakesnp_norm_var2_2 <- rnorm(10, mean = 1, sd = 0.001)
fakesnp_norm_var2_3 <- rnorm(10, mean = 1, sd = 0.001)
fakesnpvar2 <- rbind(fakesnp_norm_var2_1,
                     fakesnp_norm_var2_2,
                     fakesnp_norm_var2_3)
```

```{r results = "hide"}
fakemod_mamba2 <- mamba(betajk = fakesnpeffect2,
                        sjk2 = fakesnpvar2)
```

The outlier probabilities from MAMBA are given in the table below.
```{r echo = FALSE}
pander(fakemod_mamba2$outliermat)
```

The PPR value for the first SNP is given by `r fakemod_mamba2$ppr[1]`


```{r}
norm_prp_mod <- posterior_prp(beta = fakesnpeffect2[3 ],
                              se = sqrt(fakesnpvar2[3, ]))
norm_prp_mod$pvalue
```

In contrast, the PPR value for the same SNP is `r norm_prp_mod$pvalue`. 

\newpage
## Case 3: Normal Distribution, Zero Effect Size

In the third case, we look at SNPs with zero effect size but with differing concordant signs.
```{r}
# generating normal data centered around zero
fakesnp_norm3_1 <- rnorm(10, mean= 0, sd = 2)
fakesnp_norm3_2 <- rnorm(10, mean= 0, sd = 1)
fakesnp_norm3_3 <- rnorm(10, mean = 0, sd = 1)
fakesnpeffect3 <- rbind(fakesnp_norm3_1,
                        fakesnp_norm3_2,
                        fakesnp_norm3_3)

fakesnp_norm_var3_1 <- rnorm(10, mean = 2, sd = 0.1)
fakesnp_norm_var3_2 <- rnorm(10, mean = 1, sd = 0.1)
fakesnp_norm_var3_3 <- rnorm(10, mean = 1, sd = 0.001)
fakesnpvar3 <- rbind(fakesnp_norm_var3_1,
                     fakesnp_norm_var3_2,
                     fakesnp_norm_var3_3)
```

```{r results = "hide"}
# fitting MAMBA model
fakemod_norm3 <- mamba(betajk = fakesnpeffect3,
                       sjk2 = fakesnpvar3)
```

The outlier probabilities from MAMBA are given in the table below.
```{r echo = FALSE}
pander(fakemod_norm3$outliermat)

```

The PPR value for the first SNP is given by `r fakemod_norm3$ppr[1]`.


```{r}
# fitting posterior prp
norm_prp_mod3 <- posterior_prp(beta = fakesnpeffect3[1, ],
                               se = sqrt(fakesnpvar3[1, ]))
```

The posterior PRP value given with these SNPs is `r norm_prp_mod3$pvalue`.

\newpage
# Different Population Sampling

First we sample from 5000 individuals. We sample at different allele frequencies.
```{r}
# allele sampling
n = 5000
allele0.01 <- rbinom(n, 2, 0.01)
allele0.1 <- rbinom(n, 2, 0.1)
allele0.25 <- rbinom(n, 2, 0.25)

# noise
epsilon <- rnorm(5000, mean = 0, sd = 0.05)
```

Using a linear equation, we get our phenotype values.
```{r}
beta = 0.005
y0.01 = (beta * allele0.01) + epsilon
y0.1 = (beta * allele0.1) + epsilon
y0.25 = (beta * allele0.25) + epsilon
```

Finally we fit a linear model to get an effect size estimate.
```{r}
lm_model0.01 <- lm(y0.01 ~ allele0.01)
bhat0.01 <- lm_model0.01$coefficients[2]

lm_model0.1 <- lm(y0.1 ~ allele0.1)
bhat0.1 <- lm_model0.1$coefficients[2]

lm_model0.25 <- lm(y0.25 ~ allele0.25)
bhat0.25 <- lm_model0.25$coefficients[2]
```

The effect size estimates are given in the table below. Our true effect size is `beta`.
```{r echo = FALSE}
pander(c(bhat0.01, bhat0.1, bhat0.25))
```

Using these predictions, we can see how MAMBA and the PRP library behave.
```{r}
# list of our (predicted) effect size and variance
snpeffect <- c(bhat0.01, bhat0.1, bhat0.25)
fakesnp <- c(bhat0.01, bhat0.1, bhat0.25)
snpeffect_prp <- rbind(snpeffect, fakesnp)
snpvar <- c(var(y0.01), var(y0.1), var(y0.25))
snpvar_prp <- rbind(snpvar, snpvar)

# I doubled the values here because MAMBA seems to need more than one snp for it to work
```


First we fit to MAMBA.
```{r results = "hide"}
diff_pop <- mamba(beta = snpeffect_prp,
                  sjk2 = snpvar_prp)
```

```{r}
diff_pop$outliermat
```

Next we fit to the PRP model.

```{r}
diff_pop_prp <- posterior_prp(beta = snpeffect,
                              se = sqrt(snpvar))
```

Using MAMBA, the PPR is `r diff_pop$ppr[1]`. In contrast, using the PRP library our PRP value is `r diff_pop_prp$pvalue`.