---
title: "R Notebook, Week of 8/18"
output:
  pdf_document: default
  html_document:
    df_print: paged
  html_notebook: default
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE
)
```



```{r include = FALSE}
# libraries and other things
library(PRP)
library(rstudioapi)
library(mamba)
library(data.table)
library(parallel)
library(knitr)
library(pander)
library(ggplot2)
library(tidyverse)
library(dplyr)

#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

set.seed(2021)
```

# MAMBA Simulations
First we simulate data according to MAMBA and calculate the PPR values (from the MAMBA package) and the PRP values (from the PRP package). The job R files used for this are located in the code/data directories. The packages are calculated for all 50000 SNPS. Here, we'll look at the case where the nonoutlier study rate is 0.975.

```{r}
load(file = "data/mamba_data/sim_mamba_mod_p975.rda")
load(file = "data/mamba_data/mamba_data_p975.rda")

pprs <- sim_mod$ppr
```


We take the indices for the SNPs that have at least one outlier study.
```{r}
# indices for snps w/ and w/o outliers
out_studies <- mamba_data$Ojk
out_rows_ind <- which(rowSums(out_studies == 0) > 0) # indices of snps with outlier studies
no_out_rows_ind <-which(rowSums(out_studies) == 10) # indices of rows w/o outliers
```


Then we load in the data containing the PRPs (from PRP package) for each SNP.
```{r}
# loading PRP data

load(file = "data/prp_data/post_prp_data_pval_p975.rda") # post_prp_data_pval

```


Below are the PPRs and PRPs from the the SNPs with and without outlier studies.
```{r}
# MAMBA pprs for snps w/ and w/o outliers
out_ppr <- pprs[out_rows_ind]
nonout_ppr <- pprs[no_out_rows_ind]

# prps for snps w/ and w/o outliers
out_prp <- post_prp_data_pval[out_rows_ind]
nonout_prp <- post_prp_data_pval[no_out_rows_ind]
```

For our SNPs **with outliers**, the distribution of our PPRs looks as follows:
```{r warning = FALSE}
ggplot(data = as.data.frame(out_ppr), aes(out_ppr)) +
  geom_histogram(binwidth = 0.001) +
  ggtitle("Distribution of PPR, SNPs With Outliers") +
  xlab("PPR")
```

Rescaling the graph by ignoring some outliers, we get the histogram below.
```{r warning = FALSE}
ggplot(data = as.data.frame(out_ppr), aes(out_ppr)) +
  geom_histogram(binwidth = 0.001) +
  ggtitle("Distribution of PPR, SNPs With Outliers") +
  xlab("PPR") +
  xlim(0, 0.1)

out_mamba_sig_count <- sum(out_ppr<=0.05)
```

With outliers, there are around `r out_mamba_sig_count` SNPS with PPR values less than or equal to 0.05 (our significant SNPs).

In contrast, this is what the distribution of our PRPs for outlier SNPs looks like. Keep in mind the differently scaled x-axis.
```{r warning = FALSE}
ggplot(data = as.data.frame(out_prp), aes(out_prp)) +
  geom_histogram(binwidth = 0.001) +
  ggtitle("Distribution of PRP, SNPs With Outliers") +
  xlab("PRP")

out_prp_sig_count <- sum(out_ppr<=0.05)
```

With outliers, there are around `r out_prp_sig_count` SNPS with PPR values less than or equal to 0.05.


For our SNPs **without outliers**, the PRP distribution is given in the following histogram.

```{r warning = FALSE}
ggplot(data = as.data.frame(nonout_ppr), aes(nonout_ppr)) +
  geom_histogram(binwidth = 0.0001) +
  ggtitle("Distribution of PPR, SNPs Without Outliers") +
  xlab("PPR")
```

Like before, if we ignore extreme PPR values, our histogram looks different.
```{r warning = FALSE}
ggplot(data = as.data.frame(nonout_ppr), aes(nonout_ppr)) +
  geom_histogram(binwidth = 0.0001) +
  ggtitle("Distribution of PPR, SNPs Without Outliers") +
  xlab("PPR") +
  xlim(0, 0.1)


nout_mamba_sig_count <- sum(nonout_ppr<=0.05)
total_mamba_sig <- out_mamba_sig_count + nout_mamba_sig_count

```
The number of known nonoutlier SNPs with PPR values less than or equal to 0.05 is `r nout_mamba_sig_count`. The total number of SNPs with PPRs less than or equal to 0.05 using the MAMBA method is `r total_mamba_sig`.


In contrast, this is what the distribution of our PRP values looks like.
```{r warning = FALSE}
ggplot(data = as.data.frame(nonout_prp), aes(nonout_prp)) +
  geom_histogram(binwidth = 0.001) +
  ggtitle("Distribution of PRP, SNPs Without Outliers") +
  xlab("PRP")

nout_prp_sig_count <- sum(nonout_prp<=0.05)
total_prp_sig <- out_prp_sig_count + nout_prp_sig_count
```

The number of known nonoutlier SNPs with PPR values less than or equal to 0.05 is `r nout_mamba_sig_count`. The total number of SNPs with PPRs less than or equal to 0.05 using the MAMBA method is `r total_prp_sig`. Recall that here, our nonoutlier study rate is 0.975.


## Inverse Variance Weighted
We try an inverse variance weighted method to estimate the beta parameters and get z-score estimates.
```{r include = FALSE}
# some functions for finding the inverse variance weighted beta and se
ivw_beta <- function(snp_beta, snp_se) {
  # for one SNP
  weights <- 1/(snp_se^2)
  beta_val <- sum(snp_beta*weights)/sum(weights)
  return(beta_val)
  
}

ivw_beta_all <- function(snp_beta, snp_se) {
  # for matrix of SNPs
  beta_vals <- sapply(1:nrow(snp_beta),
                      function(row) {
                        ivw_beta(snp_beta[row, ], snp_se[row, ])
                      } )
  return(beta_vals)
}

ivw_se <- function(snp_se) {
  # for one SNP
  weights <- 1/(snp_se^2)
  sum_weights_inv <- sum(1/weights)
  beta_est_se <- sqrt(sum_weights_inv)
  return(beta_est_se)
}

ivw_se_all <- function(snp_se) {
  # for matrix of SNPs
  ivw_se_vals <- apply(snp_se,
                       MARGIN=1,
                       FUN = ivw_se)
  return(ivw_se_vals)
}
```

```{r include = FALSE}
# beta and se^2 values from mamba simulated data
beta <- mamba_data$betajk
se2 <- mamba_data$sjk2
se <- sqrt(se2)

ivw_beta_val <- ivw_beta_all(snp_beta = beta,
                               snp_se = se)
ivw_se_val <- ivw_se_all(snp_se = se)

ivw_z_scores <- ivw_beta_val/ivw_se_val

ivw_df <- data.frame(ivw_beta_val, ivw_se_val, ivw_z_scores) %>%
  rename(beta = ivw_beta_val,
         se = ivw_se_val,
         zscore = ivw_z_scores)
```

Below, we see the distribution of z-scores from inverse variance weighting.
```{r}
ggplot(ivw_df, aes(x = zscore)) +
  geom_histogram(binwidth = 0.001) +
  ggtitle("Distribution of Z-Scores from Inverse Variance Weighting") +
  xlab("Z-Scores")
  
```


\newpage
# Giant Consortium Studies
```{r include = FALSE}
get_val <- function(file_name, old_col, new_col) {
  # read in files and get values
  table_select <- read.table(file = file_name, header = TRUE) %>%
    select(c("CHR", "POS", old_col)) 
  
  for (i in 1:length(old_col)) {
    table_select <- table_select %>%
      rename(!!new_col[i] := !!old_col[i])
  }
  
  return(table_select)
}
```

## BMI
First we look at the BMI. We'll only look at the SNPs that are significant (where the p-values for "All" data file is less than 0.05/the number of SNPs).

```{r include = FALSE}
# load data if cleaned data file already exists
if (file.exists(file = "data/giant_bmi/bmi_beta.rda") &&
    file.exists(file = "data/giant_bmi/bmi_se.rda")) {
  
  load(file = "data/giant_bmi/bmi_beta.rda")
  load(file = "data/giant_bmi/bmi_se.rda")
  
  
} else {
  # height datasets
  bmi_all <- get_val(file = "data/giant_bmi/BMI_All_ancestry.fmt.txt",
                        old_col = c("beta", "se", "Pvalue"),
                        new_col = c("beta_all", "se_all", "pval"))
  bmi_AA <- get_val(file = "data/giant_bmi/BMI_African_American.fmt.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_AA", "se_AA"))
  bmi_EA <- get_val(file = "data/giant_bmi/BMI_Eastern_Asian.fmt.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_EA", "se_EA"))
  bmi_EU <- get_val(file = "data/giant_bmi/BMI_European.fmt.txt",
                        old_col = c("beta", "se"),
                        new_col = c("beta_EAS", "se_EAS"))
  bmi_HA <- get_val(file = "data/giant_bmi/BMI_Hispanic_American.fmt.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_HA", "se_HA"))
  bmi_SA <- get_val(file = "data/giant_bmi/BMI_South_Asian.fmt.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_SA", "se_SA"))
  
  
  # put all values together
  bmi_list <- list(bmi_AA, bmi_EA, bmi_EU, bmi_HA, bmi_SA, bmi_all)
  
  
  # merging data and dropping chr and pos columns
  bmi_vals <- Reduce(function(df1, df2) {
    merge(x = df1, y = df2, by = c("CHR", "POS"), all = TRUE)
  },
  bmi_list) %>%
    filter(pval <= 0.05/nrow(bmi_all))
  
  
  # remove any rows that are all NA
  bmi_vals[bmi_vals == Inf] <- NA
  
  
  # indices that are all NA values
  allnaind <- bmi_vals %>%
    select(-c("CHR", "POS", "beta_all", "se_all", "pval")) %>%
    apply(MARGIN=1, function(row_val) {
      all(is.na(row_val))
    } )
  
  bmi_vals <- bmi_vals[!allnaind, ] # exclude rows that are all NA
  
  
  # extracting beta and se and saving it
  beta_cols <- c("beta_AA", "beta_EA", "beta_EAS", "beta_HA", "beta_SA")
  bmi_beta <- bmi_vals %>%
    select(beta_cols)
  
  se_cols <- c("se_AA", "se_EA", "se_EAS", "se_HA", "se_SA")
  bmi_se <- bmi_vals %>%
    select(se_cols)
  
  
  save(bmi_beta, file = "data/giant_bmi/bmi_beta.rda")
  save(bmi_se, file = "data/giant_bmi/bmi_se.rda")
  
  load(file = "data/giant_bmi/bmi_beta.rda")
  load(file = "data/giant_bmi/bmi_se.rda")
  
}
```


### BMI Graphs
Now we get a better look at the BMI through graphical representations.

```{r include = FALSE}
source(file = "data/giant_replicable.R")
```


```{r include = FALSE}
if (file.exists(file = "data/giant_bmi/bmi_prp.rda") &&
    file.exists(file = "data/giant_bmi/bmi_prp_pval.rda") &&
    file.exists(file = "data/giant_bmi/bmi_mamba.rda")) {
  
  load(file = "data/giant_bmi/bmi_prp.rda")
  load(file = "data/giant_bmi/bmi_prp_pval.rda")
  load(file = "data/giant_bmi/bmi_mamba.rda")
  
} else {
  
  giant_replicable(beta = bmi_beta,
                   se = bmi_se,
                   prp_file = "data/giant_bmi/bmi_prp.rda",
                   prp_pval_file = "data/giant_bmi/bmi_prp_pval.rda",
                   mamba_data_file = "data/giant_bmi/bmi_mamba.rda")
  
  load(file = "data/giant_bmi/bmi_prp.rda")
  load(file = "data/giant_bmi/bmi_prp_pval.rda")
  load(file = "data/giant_bmi/bmi_mamba.rda")
  
}

```

We take a look at our PPR and PRP values. The histogram of our PPR values for BMI is given below.

```{r echo = FALSE, message = FALSE}
# from MAMBA
bmi_ppr_val <- as.data.frame(mamba_data$ppr)
ggplot(data = bmi_ppr_val, aes(x = mamba_data$ppr)) +
  geom_histogram(binwidth = 0.01) +
  xlab("MAMBA PPR Values")
```

In contrast, here is our histogram of PRP values.

```{r echo = FALSE, message = FALSE}
# from PRP
bmi_prp_pval <- as.data.frame(post_prp_data_pval) %>%
  rename("prp_pval" = "post_prp_data_pval")

ggplot(data = bmi_prp_pval, aes(x = prp_pval)) +
  geom_histogram(binwidth = 0.01) +
  xlab("PRP Values")
```


\newpage
## Height
Next we look at height.

```{r include = FALSE}
# loading in cleaned data, or if it doesn't exist then cleaning it now
if (file.exists(file = "data/giant_height/height_beta.rda") &&
    file.exists(file = "data/giant_height/height_se.rda")) {
  
  load(file = "data/giant_height/height_beta.rda")
  load(file = "data/giant_height/height_se.rda")
  
} else {
  
  # height datasets
  height_all <- get_val(file = "data/giant_height/height_All_add_SV.txt",
                        old_col = c("beta", "se", "Pvalue"),
                        new_col = c("beta_all", "se_all", "pval"))
  height_AA <- get_val(file = "data/giant_height/height_AA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_AA", "se_AA"))
  height_EA <- get_val(file = "data/giant_height/height_EA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_EA", "se_EA"))
  height_EAS <- get_val(file = "data/giant_height/height_EAS_add_SV.txt",
                        old_col = c("beta", "se"),
                        new_col = c("beta_EAS", "se_EAS"))
  height_HA <- get_val(file = "data/giant_height/height_HA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_HA", "se_HA"))
  height_SA <- get_val(file = "data/giant_height/height_SA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_SA", "se_SA"))
  
  
  # put all values together
  height_list <- list(height_AA, height_EA, height_EAS, height_HA, height_SA, height_all)
  
  
  # merging data and dropping chr and pos columns
  height_vals <- Reduce(function(df1, df2) {
    merge(x = df1, y = df2, by = c("CHR", "POS"), all = TRUE)
  },
  height_list) %>%
    filter(pval <= 0.05/nrow(height_all))
  
  
  # remove any rows that are all NA
  height_vals[height_vals == Inf] <- NA
  
  
  # indices that are all NA values
  allnaind <- height_vals %>%
    select(-c("CHR", "POS", "beta_all", "se_all", "pval")) %>%
    apply(MARGIN=1, function(row_val) {
      all(is.na(row_val))
    } )
  
  height_vals <- height_vals[!allnaind, ] # exclude rows that are all NA
  
  
  # extracting beta and se and saving it
  beta_cols <- c("beta_AA", "beta_EA", "beta_EAS", "beta_HA", "beta_SA")
  height_beta <- height_vals %>%
    select(beta_cols)
  
  se_cols <- c("se_AA", "se_EA", "se_EAS", "se_HA", "se_SA")
  height_se <- height_vals %>%
    select(se_cols)
  
  
  save(height_beta, file = "data/giant_height/height_beta.rda")
  save(height_se, file = "data/giant_height/height_se.rda")  
  
}
```

```{r include = FALSE}
if (file.exists(file = "data/giant_height/height_prp.rda") &&
    file.exists(file = "data/giant_height/height_prp_val.rda") &&
    file.exists(file = "data/giant_height/height_mamba.rda")) {
  
  load(file = "data/giant_height/height_prp.rda")
  load(file = "data/giant_height/height_prp_val.rda")
  load(file = "data/giant_height/height_mamba.rda")
  
} else {
  
  giant_replicable(beta = height_beta,
                   se = height_se,
                   prp_file = "data/giant_height/height_prp.rda",
                   prp_pval_file = "data/giant_height/height_prp_val.rda",
                   mamba_data_file = "data/giant_height/height_mamba.rda")
  
  load(file = "data/giant_height/height_prp.rda")
  load(file = "data/giant_height/height_prp_val.rda")
  load(file = "data/giant_height/height_mamba.rda")
  
}


```

### Height Graphs
We take a look at our PPR and PRP values. The histogram of our PPR values is given below.

```{r echo = FALSE, message = FALSE}
# from MAMBA
height_ppr_val <- as.data.frame(mamba_data$ppr)
ggplot(data = height_ppr_val, aes(x = mamba_data$ppr)) +
  geom_histogram() +
  xlab("MAMBA PPR Values")
```

In contrast, we have our histogram for PRP values.

```{r echo = FALSE, message = FALSE}
# from PRP
height_prp_pval <- as.data.frame(post_prp_data_pval) %>%
  rename("prp_pval" = "post_prp_data_pval")

ggplot(data = height_prp_pval, aes(x = prp_pval)) +
  geom_histogram() +
  xlab("PRP Values")
```


# Prior Checking

```{r include = FALSE}
# data cleaning
if (file.exists(file = "data/giant_meta/meta_beta.rda") &&
    file.exists(file = "data/giant_meta/meta_se.rda")) {
  
  load(file = "data/giant_meta/meta_beta.rda")
  load(file = "data/giant_meta/meta_se.rda")
  
} else {
  # read in data files
  old_cols <- c("SNP", "BETA", "SE", "P")
  
  new_cols_locke <- c("snp", "beta_locke", "se_locke", "pval_locke")
  meta_Locke <- get_val(file_name = "data/giant_meta/Meta-analysis_Locke_et_al+UKBiobank_2018_UPDATED.txt",
                        old_col = old_cols,
                        new_col = new_cols_locke)
  
  new_cols_wood <- c("snp", "beta_wood", "se_wood", "pval_wood")
  meta_Wood <- get_val(file_name = "data/giant_meta/Meta-analysis_Wood_et_al+UKBiobank_2018.txt",
                       old_col = old_cols,
                       new_col = new_cols_wood)
  
  # combining data
  meta_df <- merge(x = meta_Locke,
                   y = meta_Wood,
                   by = c("CHR", "POS", "snp"),
                   all = TRUE)
  
  # filtering to keep significant snps
  meta_df <- filter(meta_df, pval_locke <= 0.05/nrow(meta_df))
  
  
  # remove any rows that are all NA
  meta_df[meta_df == Inf] <- NA
  
  
  # indices that are all NA values
  allnaind <- meta_df %>%
    select(-c("CHR", "POS", "snp", "pval_locke", "pval_wood")) %>%
    apply(MARGIN=1, function(row_val) {
      all(is.na(row_val))
    } )
  
  meta_df <- meta_df[!allnaind, ]
  
  
  meta_beta <- meta_df %>%
    select(c("beta_locke", "beta_wood"))
  
  meta_se <- meta_df %>%
    select(c("se_locke", "se_wood"))
  
  
  save(meta_beta, file = "data/giant_meta/meta_beta.rda")
  save(meta_se, file = "data/giant_meta/meta_se.rda")
}
```

```{r include = FALSE}
load("data/giant_meta/meta_prp_file.rda")
load("data/giant_meta/meta_prp_pval.rda")
load("data/giant_meta/meta_mamba.rda")
```

## Meta Graphs
We take a look at our PPR and PRP values. The histogram of our PPR values is given below.

```{r echo = FALSE, message = FALSE}
# from MAMBA
meta_ppr_val <- as.data.frame(mamba_data$ppr)
ggplot(data = meta_ppr_val, aes(x = mamba_data$ppr)) +
  geom_histogram() +
  xlab("Meta PPR Values")
```

In contrast, we have our histogram for PRP values.

```{r echo = FALSE, message = FALSE}
# from PRP
meta_prp_pval <- as.data.frame(prior_prp_data_pval) %>%
  rename("prp_pval" = "prior_prp_data_pval")

ggplot(data = meta_prp_pval, aes(x = prp_pval)) +
  geom_histogram() +
  xlab("PRP Values")
```