---
title: "R Notebook, Week of 8/18"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE
)
```



```{r include = FALSE}
# libraries and other things
library(PRP)
library(rstudioapi)
library(mamba)
library(data.table)
library(parallel)
library(knitr)
library(pander)
library(ggplot2)
library(tidyverse)
library(dplyr)

#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

set.seed(2021)
```


First we simulate data according to MAMBA and calculate the PPR values (from the MAMBA package) and the PRP values (from the PRP package). The job R files used for this are located in the code/data directories. The packages are calculated for all 50000 SNPS. Here, we'll look at the case where the nonoutlier study rate is 0.975.

```{r}
load(file = "data/mamba_data/sim_mamba_mod_p975.rda")
load(file = "data/mamba_data/mamba_data_p975.rda")

pprs <- sim_mod$ppr
```


We take the indices for the SNPs that have at least one outlier study.
```{r}
# indices for snps w/ and w/o outliers
out_studies <- mamba_data$Ojk
out_rows_ind <- which(rowSums(out_studies == 0) > 0) # indices of snps with outlier studies
no_out_rows_ind <-which(rowSums(out_studies) == 10) # indices of rows w/o outliers
```


Then we load in the data containing the PRPs (from PRP package) for each SNP.
```{r}
# loading PRP data

load(file = "data/prp_data/post_prp_data_pval_p975.rda") # post_prp_data_pval

```


Below are the PPRs and PRPs from the the SNPs with and without outlier studies.
```{r}
# MAMBA pprs for snps w/ and w/o outliers
out_ppr <- pprs[out_rows_ind]
nonout_ppr <- pprs[no_out_rows_ind]

# prps for snps w/ and w/o outliers
out_prp <- post_prp_data_pval[out_rows_ind]
nonout_prp <- post_prp_data_pval[no_out_rows_ind]
```

For our SNPs **with outliers**, the distribution of our PPRs looks as follows:
```{r warning = FALSE}
ggplot(data = as.data.frame(out_ppr), aes(out_ppr)) +
  geom_histogram(binwidth = 0.001) +
  ggtitle("Distribution of PPR, SNPs With Outliers") +
  xlab("PPR")
```

Rescaling the graph by ignoring some outliers, we get the histogram below.
```{r warning = FALSE}
ggplot(data = as.data.frame(out_ppr), aes(out_ppr)) +
  geom_histogram(binwidth = 0.001) +
  ggtitle("Distribution of PPR, SNPs With Outliers") +
  xlab("PPR") +
  xlim(0, 0.1)
```

In contrast, this is what the distribution of our PRPs for outlier SNPs looks like. Keep in mind the differently scaled x-axis.
```{r warning = FALSE}
ggplot(data = as.data.frame(out_prp), aes(out_prp)) +
  geom_histogram(binwidth = 0.001) +
  ggtitle("Distribution of PRP, SNPs With Outliers") +
  xlab("PRP")
```


For our SNPs **without outliers**, the PRP distribution is given in the following histogram.

```{r warning = FALSE}
ggplot(data = as.data.frame(nonout_ppr), aes(nonout_ppr)) +
  geom_histogram(binwidth = 0.0001) +
  ggtitle("Distribution of PPR, SNPs Without Outliers") +
  xlab("PPR")
```

Like before, if we ignore extreme PPR values, our histogram looks different.
```{r warning = FALSE}
ggplot(data = as.data.frame(nonout_ppr), aes(nonout_ppr)) +
  geom_histogram(binwidth = 0.0001) +
  ggtitle("Distribution of PPR, SNPs Without Outliers") +
  xlab("PPR") +
  xlim(0, 0.1)
```

In contrast, this is what the distribution of our PRP values looks like.
```{r warning = FALSE}
ggplot(data = as.data.frame(nonout_prp), aes(nonout_prp)) +
  geom_histogram(binwidth = 0.001) +
  ggtitle("Distribution of PRP, SNPs Without Outliers") +
  xlab("PRP")
```

\newpage
# Giant Consortium Studies
```{r include = FALSE}
get_val <- function(file_name, old_col, new_col) {
  # read in files and get values
  table_select <- read.table(file = file_name, header = TRUE) %>%
    select(c("CHR", "POS", old_col)) 
  
  for (i in 1:length(old_col)) {
    table_select <- table_select %>%
      rename(!!new_col[i] := !!old_col[i])
  }
  
  return(table_select)
}
```

## BMI
First we look at the BMI. We'll only look at the SNPs that are significant (where the p-values for "All" data file is less than 0.05/the number of SNPs).

```{r include = FALSE}
# load data if cleaned data file already exists
if (file.exists(file = "data/giant_bmi/bmi_beta.rda") &&
    file.exists(file = "data/giant_bmi/bmi_se.rda")) {
  
  load(file = "data/giant_bmi/bmi_beta.rda")
  load(file = "data/giant_bmi/bmi_se.rda")
  
  
} else {
  # height datasets
  bmi_all <- get_val(file = "data/giant_bmi/BMI_All_ancestry.fmt.txt",
                        old_col = c("beta", "se", "Pvalue"),
                        new_col = c("beta_all", "se_all", "pval"))
  bmi_AA <- get_val(file = "data/giant_bmi/BMI_African_American.fmt.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_AA", "se_AA"))
  bmi_EA <- get_val(file = "data/giant_bmi/BMI_Eastern_Asian.fmt.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_EA", "se_EA"))
  bmi_EU <- get_val(file = "data/giant_bmi/BMI_European.fmt.txt",
                        old_col = c("beta", "se"),
                        new_col = c("beta_EAS", "se_EAS"))
  bmi_HA <- get_val(file = "data/giant_bmi/BMI_Hispanic_American.fmt.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_HA", "se_HA"))
  bmi_SA <- get_val(file = "data/giant_bmi/BMI_South_Asian.fmt.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_SA", "se_SA"))
  
  
  # put all values together
  bmi_list <- list(bmi_AA, bmi_EA, bmi_EU, bmi_HA, bmi_SA, bmi_all)
  
  
  # merging data and dropping chr and pos columns
  bmi_vals <- Reduce(function(df1, df2) {
    merge(x = df1, y = df2, by = c("CHR", "POS"), all = TRUE)
  },
  bmi_list) %>%
    filter(pval <= 0.05/nrow(bmi_all))
  
  
  # remove any rows that are all NA
  bmi_vals[bmi_vals == Inf] <- NA
  
  
  # indices that are all NA values
  allnaind <- bmi_vals %>%
    select(-c("CHR", "POS")) %>%
    apply(MARGIN=1, function(row_val) {
      all(is.na(row_val))
    } )
  
  bmi_vals <- bmi_vals[!allnaind, ] # exclude rows that are all NA
  
  
  # extracting beta and se and saving it
  beta_cols <- c("beta_AA", "beta_EA", "beta_EAS", "beta_HA", "beta_SA")
  bmi_beta <- bmi_vals %>%
    select(beta_cols)
  
  se_cols <- c("se_AA", "se_EA", "se_EAS", "se_HA", "se_SA")
  bmi_se <- bmi_vals %>%
    select(se_cols)
  
  
  save(bmi_beta, file = "data/giant_bmi/bmi_beta.rda")
  save(bmi_se, file = "data/giant_bmi/bmi_se.rda")
  
  
}
```


## BMI Graphs
Now we get a better look at the BMI through graphical representations.

```{r include = FALSE}
source(file = "data/giant_replicable.R")
```


```{r include = FALSE}
if (file.exists(file = "data/giant_bmi/bmi_prp.rda") &&
    file.exists(file = "data/giant_bmi/bmi_prp_pval.rda") &&
    file.exists(file = "data/giant_bmi/bmi_mamba.rda")) {
  
  load(file = "data/giant_bmi/bmi_prp.rda")
  load(file = "data/giant_bmi/bmi_prp_pval.rda")
  load(file = "data/giant_bmi/bmi_mamba.rda")
  
} else {
  
  giant_replicable(beta = bmi_beta,
                   se = bmi_se,
                   prp_file = "data/giant_bmi/bmi_prp.rda",
                   prp_pval_file = "data/giant_bmi/bmi_prp_pval.rda",
                   mamba_data_file = "data/giant_bmi/bmi_mamba.rda")
  
  load(file = "data/giant_bmi/bmi_prp.rda")
  load(file = "data/giant_bmi/bmi_prp_pval.rda")
  load(file = "data/giant_bmi/bmi_mamba.rda")
  
}

```

We take a look at our PPR and PRP values. The histogram of our PPR values for BMI is given below.

```{r echo = FALSE, message = FALSE}
# from MAMBA
bmi_ppr_val <- as.data.frame(mamba_data$ppr)
ggplot(data = bmi_ppr_val, aes(x = mamba_data$ppr)) +
  geom_histogram() +
  xlab("MAMBA PPR Values")
```

In contrast, here is our histogram of PRP values.

```{r echo = FALSE, message = FALSE}
# from PRP
bmi_prp_pval <- as.data.frame(post_prp_data_pval) %>%
  rename("prp_pval" = "post_prp_data_pval")

ggplot(data = bmi_prp_pval, aes(x = prp_pval)) +
  geom_histogram() +
  xlab("PRP Values")
```


\newpage
# Height
Next we look at height.

```{r include = FALSE}
# loading in cleaned data, or if it doesn't exist then cleaning it now
if (file.exists(file = "data/giant_height/height_beta.rda") &&
    file.exists(file = "data/giant_height/height_se.rda")) {
  
  load(file = "data/giant_height/height_beta.rda")
  load(file = "data/giant_height/height_se.rda")
  
} else {
  
  # height datasets
  height_all <- get_val(file = "data/giant_height/height_All_add_SV.txt",
                        old_col = c("beta", "se", "Pvalue"),
                        new_col = c("beta_all", "se_all", "pval"))
  height_AA <- get_val(file = "data/giant_height/height_AA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_AA", "se_AA"))
  height_EA <- get_val(file = "data/giant_height/height_EA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_EA", "se_EA"))
  height_EAS <- get_val(file = "data/giant_height/height_EAS_add_SV.txt",
                        old_col = c("beta", "se"),
                        new_col = c("beta_EAS", "se_EAS"))
  height_HA <- get_val(file = "data/giant_height/height_HA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_HA", "se_HA"))
  height_SA <- get_val(file = "data/giant_height/height_SA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_SA", "se_SA"))
  
  
  # put all values together
  height_list <- list(height_AA, height_EA, height_EAS, height_HA, height_SA, height_all)
  
  
  # merging data and dropping chr and pos columns
  height_vals <- Reduce(function(df1, df2) {
    merge(x = df1, y = df2, by = c("CHR", "POS"), all = TRUE)
  },
  height_list) %>%
    filter(pval <= 0.05/nrow(height_all))
  
  
  # remove any rows that are all NA
  height_vals[height_vals == Inf] <- NA
  
  
  # indices that are all NA values
  allnaind <- height_vals %>%
    select(-c("CHR", "POS")) %>%
    apply(MARGIN=1, function(row_val) {
      all(is.na(row_val))
    } )
  
  height_vals <- height_vals[!allnaind, ] # exclude rows that are all NA
  
  
  # extracting beta and se and saving it
  beta_cols <- c("beta_AA", "beta_EA", "beta_EAS", "beta_HA", "beta_SA")
  height_beta <- height_vals %>%
    select(beta_cols)
  
  se_cols <- c("se_AA", "se_EA", "se_EAS", "se_HA", "se_SA")
  height_se <- height_vals %>%
    select(se_cols)
  
  
  save(height_beta, file = "data/giant_height/height_beta.rda")
  save(height_se, file = "data/giant_height/height_se.rda")  
  
}
```

```{r include = FALSE}
if (file.exists(file = "data/giant_height/height_prp.rda") &&
    file.exists(file = "data/giant_height/height_prp_val.rda") &&
    file.exists(file = "data/giant_height/height_mamba.rda")) {
  
  load(file = "data/giant_height/height_prp.rda")
  load(file = "data/giant_height/height_prp_val.rda")
  load(file = "data/giant_height/height_mamba.rda")
  
} else {
  
  giant_replicable(beta = height_beta,
                   se = height_se,
                   prp_file = "data/giant_height/height_prp.rda",
                   prp_pval_file = "data/giant_height/height_prp_val.rda",
                   mamba_data_file = "data/giant_height/height_mamba.rda")
  
  load(file = "data/giant_height/height_prp.rda")
  load(file = "data/giant_height/height_prp_val.rda")
  load(file = "data/giant_height/height_mamba.rda")
  
}


```

We take a look at our PPR and PRP values. The histogram of our PPR values is given below.

```{r echo = FALSE, message = FALSE}
# from MAMBA
height_ppr_val <- as.data.frame(mamba_data$ppr)
ggplot(data = height_ppr_val, aes(x = mamba_data$ppr)) +
  geom_histogram() +
  xlab("MAMBA PPR Values")
```

In contrast, we have our histogram for PRP values.

```{r echo = FALSE, message = FALSE}
# from PRP
height_prp_pval <- as.data.frame(post_prp_data_pval) %>%
  rename("prp_pval" = "post_prp_data_pval")

ggplot(data = height_prp_pval, aes(x = prp_pval)) +
  geom_histogram() +
  xlab("PRP Values")
```

