---
title: "R Notebook, Week of 8/18"
output:
  html_document:
    df_print: paged
    toc: true
  pdf_document:
    toc: true
  html_notebook: default
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE
)
```


\newpage
```{r Libraries, include = FALSE}
# libraries and other things
library(PRP)
library(rstudioapi)
library(mamba)
library(data.table)
library(parallel)
library(knitr)
library(pander)
library(ggplot2)
library(tidyverse)
library(dplyr)

#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

set.seed(2021)
```
# Background


\newpage
# MAMBA Simulations
First we simulate data according to MAMBA and calculate the PPR values (from the MAMBA package) and the PRP values (from the PRP package). The job R files used for this are located in the code/data directories. The packages are calculated for all 50000 SNPS. Here, we'll look at the case where the nonoutlier study rate for the generated MAMBA simulations is 0.975.

```{r}
# loading mamba and prp data
load(file = "data/mamba_data/sim_mamba_mod_p975.rda")
load(file = "data/mamba_data/mamba_data_p975.rda")
load(file = "data/prp_data/post_prp_data_pval_p975.rda") # post_prp_data_pval

pprs <- sim_mod$ppr
```


We take the indices for the SNPs that have at least one outlier study. We use these indices to compare against the results of the MAMBA and PRP libraries.
```{r}
# indices for snps w/ and w/o outliers
out_studies <- mamba_data$Ojk
out_rows_ind <- which(rowSums(out_studies == 0) > 0) # indices of snps with outlier studies
no_out_rows_ind <-which(rowSums(out_studies) == 10) # indices of rows w/o outliers

# MAMBA pprs for snps w/ and w/o outliers
out_ppr <- pprs[out_rows_ind]
nonout_ppr <- pprs[no_out_rows_ind]

# prps for snps w/ and w/o outliers
out_prp <- post_prp_data_pval[out_rows_ind]
nonout_prp <- post_prp_data_pval[no_out_rows_ind]
```

\newpage
## SNPs With Outliers
For our SNPs **with outliers**, the distribution of our PPR values from MAMBA is given below. We can see that our distribution is heavily right skewed, with most PPR values below 0.1.
```{r, warning = FALSE}
ggplot(data = as.data.frame(out_ppr), aes(out_ppr)) +
  geom_histogram(binwidth = 0.001) +
  ggtitle("Distribution of PPR, SNPs With Outliers") +
  xlab("PPR")
```

\newpage
Rescaling the graph by ignoring some outliers, we get the histogram below. We can see it is still heavily right skewed.
```{r, warning = FALSE}
ggplot(data = as.data.frame(out_ppr), aes(out_ppr)) +
  geom_histogram(binwidth = 0.001) +
  ggtitle("Distribution of PPR, SNPs With Outliers") +
  xlab("PPR") +
  xlim(0, 0.1)
```

```{r, include = FALSE}
out_mamba_sig_count <- sum(out_ppr<=0.05)
out_mamba_nonsig_count <- length(out_ppr) - out_mamba_sig_count
```

With outliers, there are around `r out_mamba_sig_count` SNPS with PPR values less than or equal to 0.05 (our significant SNPs).

\newpage
In contrast, this is what the distribution of our PRPs for outlier SNPs looks like. Keep in mind the differently scaled x-axis.
```{r, warning = FALSE}
ggplot(data = as.data.frame(out_prp), aes(out_prp)) +
  geom_histogram(binwidth = 0.001) +
  ggtitle("Distribution of PRP, SNPs With Outliers") +
  xlab("PRP")
```

```{r, include = FALSE}
out_prp_sig_count <- sum(out_prp<=0.05)
out_prp_nonsig_count <- length(out_prp) - out_prp_sig_count
```

With outliers, there are around `r out_prp_sig_count` SNPS with PPR values less than or equal to 0.05.


\newpage
## SNPs Without Outliers
For our SNPs **without outliers**, the PRP distribution is given in the following histogram.

```{r, warning = FALSE}
ggplot(data = as.data.frame(nonout_ppr), aes(nonout_ppr)) +
  geom_histogram(binwidth = 0.0001) +
  ggtitle("Distribution of PPR, SNPs Without Outliers") +
  xlab("PPR")
```

\newpage
Like before, if we ignore extreme PPR values, we can see that much of our PPR values are below 0.025.
```{r, warning = FALSE}
ggplot(data = as.data.frame(nonout_ppr), aes(nonout_ppr)) +
  geom_histogram(binwidth = 0.0001) +
  ggtitle("Distribution of PPR, SNPs Without Outliers") +
  xlab("PPR") +
  xlim(0, 0.1)
```

```{r, include = FALSE}
nout_mamba_sig_count <- sum(nonout_ppr<=0.05)
nout_mamba_nonsig_count <- length(nonout_ppr) - nout_mamba_sig_count
total_mamba_sig <- out_mamba_sig_count + nout_mamba_sig_count

```
The number of known nonoutlier SNPs with PPR values less than or equal to 0.05 is `r nout_mamba_sig_count`. The total number of SNPs with PPRs less than or equal to 0.05 using the MAMBA method is `r total_mamba_sig`.

\newpage
In contrast, this is what the distribution of our PRP values looks like. We can see that it is much more uniform compared to the other graphs above.
```{r, warning = FALSE}
ggplot(data = as.data.frame(nonout_prp), aes(nonout_prp)) +
  geom_histogram(binwidth = 0.001) +
  ggtitle("Distribution of PRP, SNPs Without Outliers") +
  xlab("PRP")
```

```{r, include = FALSE}
nout_prp_sig_count <- sum(nonout_prp<=0.05)
nout_prp_nonsig_count <- length(nonout_prp) - nout_prp_sig_count
total_prp_sig <- out_prp_sig_count + nout_prp_sig_count
```

The number of known nonoutlier SNPs with PPR values less than or equal to 0.05 is `r nout_mamba_sig_count`. The total number of SNPs with PPRs less than or equal to 0.05 using the MAMBA method is `r total_prp_sig`. Recall that here, our nonoutlier study rate is 0.975.

\newpage
## SNP Counts
A table of the number of **outlier** SNPs that are considered nonreplicable (a PRP or PPR value of 0.05 or less) is given below for each replicability method.
```{r, echo = FALSE}
out_df <- data.frame(c(out_mamba_sig_count, out_mamba_nonsig_count),
                     c(out_prp_sig_count, out_prp_nonsig_count),
                     row.names = c("Significant", "Non-Significant"))
names(out_df) <- c("MAMBA Outlier", "PRP Outlier")

pander(out_df)
```

For comparison, a table of the number of **nonoutlier** SNPs that are considered nonreplicable is below.
```{r, echo = FALSE}
nout_df <- data.frame(c(nout_mamba_sig_count, nout_mamba_nonsig_count),
                      c(nout_prp_sig_count, nout_prp_nonsig_count),
                      row.names = c("Significant", "Non-Significant"))
names(nout_df) <- c("MAMBA Nonoutlier", "PRP Nonoutlier")
combined_table <- cbind(out_df, nout_df)
pander(nout_df)
```


A table of the total significant and nonsignicant SNPs is given below. For reference, there were a total of `r nrow(out_studies)` SNPs in the simulated data.
```{r, echo = FALSE}
total_mamba_nonsig <- nrow(out_studies) - total_mamba_sig
total_prp_nonsig <- nrow(out_studies) - total_prp_sig

total_df <- data.frame(c(total_mamba_sig, total_mamba_nonsig),
                       c(total_prp_sig, total_prp_nonsig),
                       row.names = c("Total Significant SNPs", "Total Nonsignificant SNPs"))

names(total_df) <- c("MAMBA", "PRP")

pander(total_df)
```

\newpage
# Giant Consortium Studies
```{r Replicability Function, include = FALSE}
source(file = "data/giant_replicable.R")
source(file = "data/Inv_Var_Weight.R")
```

```{r Reading Data Function, include = FALSE}
get_val <- function(file_name, old_col, new_col) {
  # read in files and get values
  table_select <- read.table(file = file_name, header = TRUE) %>%
    select(c("CHR", "POS", old_col)) 
  
  for (i in 1:length(old_col)) {
    table_select <- table_select %>%
      rename(!!new_col[i] := !!old_col[i])
  }
  
  return(table_select)
}
```

\newpage
## Height
We only look at height for now.

```{r, include = FALSE}
# loading in cleaned data, or if it doesn't exist then cleaning it now
if (file.exists(file = "data/giant_height/height_beta.rda") &&
    file.exists(file = "data/giant_height/height_se.rda")) {
  
  load(file = "data/giant_height/height_beta.rda")
  load(file = "data/giant_height/height_se.rda")
  
} else {
  
  # height datasets
  height_all <- get_val(file = "data/giant_height/height_All_add_SV.txt",
                        old_col = c("beta", "se", "Pvalue"),
                        new_col = c("beta_all", "se_all", "pval"))
  height_AA <- get_val(file = "data/giant_height/height_AA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_AA", "se_AA"))
  height_EA <- get_val(file = "data/giant_height/height_EA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_EA", "se_EA"))
  height_EAS <- get_val(file = "data/giant_height/height_EAS_add_SV.txt",
                        old_col = c("beta", "se"),
                        new_col = c("beta_EAS", "se_EAS"))
  height_HA <- get_val(file = "data/giant_height/height_HA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_HA", "se_HA"))
  height_SA <- get_val(file = "data/giant_height/height_SA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_SA", "se_SA"))
  
  
  # put all values together
  height_list <- list(height_AA, height_EA, height_EAS, height_HA, height_SA, height_all)
  
  
  # merging data and dropping chr and pos columns
  height_vals <- Reduce(function(df1, df2) {
    merge(x = df1, y = df2, by = c("CHR", "POS"), all = TRUE)
  },
  height_list) %>%
    filter(pval <= 0.05/nrow(height_all))
  
  
  # remove any rows that are all NA
  height_vals[height_vals == Inf] <- NA
  
  
  # indices that are all NA values
  allnaind <- height_vals %>%
    select(-c("CHR", "POS", "beta_all", "se_all", "pval")) %>%
    apply(MARGIN=1, function(row_val) {
      all(is.na(row_val))
    } )
  
  height_vals <- height_vals[!allnaind, ] # exclude rows that are all NA
  
  
  # extracting beta and se and saving it
  beta_cols <- c("beta_AA", "beta_EA", "beta_EAS", "beta_HA", "beta_SA")
  height_beta <- height_vals %>%
    select("CHR", "POS", beta_cols)
  
  se_cols <- c("se_AA", "se_EA", "se_EAS", "se_HA", "se_SA")
  height_se <- height_vals %>%
    select("CHR", "POS", se_cols)
  
  
  save(height_beta, file = "data/giant_height/height_beta.rda")
  save(height_se, file = "data/giant_height/height_se.rda")  
  
}
```

```{r, include = FALSE}
if (file.exists(file = "data/giant_height/height_prp.rda") &&
    file.exists(file = "data/giant_height/height_prp_val.rda") &&
    file.exists(file = "data/giant_height/height_mamba.rda")) {
  
  load(file = "data/giant_height/height_prp.rda")
  load(file = "data/giant_height/height_prp_val.rda")
  load(file = "data/giant_height/height_mamba.rda")
  
} else {
  beta_cols <- c("beta_AA", "beta_EA", "beta_EAS", "beta_HA", "beta_SA")
  height_beta_val <- height_beta %>%
    select(beta_cols)
  
  se_cols <- c("se_AA", "se_EA", "se_EAS", "se_HA", "se_SA")
  height_se_val <- height_vals %>%
    select(se_cols)
  
  giant_replicable(beta = height_beta_val,
                   se = height_se_val,
                   prp_file = "data/giant_height/height_prp.rda",
                   prp_pval_file = "data/giant_height/height_prp_val.rda",
                   mamba_data_file = "data/giant_height/height_mamba.rda")
  
  load(file = "data/giant_height/height_prp.rda")
  load(file = "data/giant_height/height_prp_val.rda")
  load(file = "data/giant_height/height_mamba.rda")
  
}


```

### Height Graphs
We take a look at our PPR and PRP values. The histogram of our PPR values is given below.

```{r, echo = FALSE, message = FALSE}
# from MAMBA
height_ppr_val <- as.data.frame(mamba_data$ppr) %>%
  rename("MAMBA" = "mamba_data$ppr")
ggplot(data = height_ppr_val, aes(x = mamba_data$ppr)) +
  geom_histogram() +
  xlab("MAMBA PPR Values")
```

We can see that the distribution of the PPR values is skewed left. Many values relatively high, with only a few that seem significant. 

In contrast, our histogram for PRP values seems more evenly distributed, with a slight left right. More values are considered significant.

```{r, echo = FALSE, message = FALSE}
# from PRP
height_prp_pval <- as.data.frame(post_prp_data_pval) %>%
  rename("PRP" = "post_prp_data_pval")

ggplot(data = height_prp_pval, aes(x = PRP)) +
  geom_histogram() +
  xlab("PRP Values")
```


```{r Height Gene Replicability Filter, include = FALSE}
# combined data
chrom_id <- height_beta %>%
  select("CHR", "POS")

# merge all data back together (IDs, )
height_rep_df <- cbind(chrom_id, height_ppr_val, height_prp_pval) %>%
  merge(y = height_beta, by = c("CHR", "POS"), all = TRUE) %>%
  merge(y = height_se, by = c("CHR", "POS"), all = TRUE)

# filter out values less than 0.05
# SNPs with MAMBA PPR and PRP values less than 0.05
sig_gene <- height_rep_df %>%
  filter(MAMBA <= 0.05 & PRP <= 0.05)

# filter, keep nonsignificant SNPs
non_sig_gene <- height_rep_df %>%
  filter(MAMBA > 0.05 | PRP > 0.05)
```


Now we'll look at how the effect sizes look between SNPs that MAMBA and PRP have deemed significant and not signficant. Below, we have a forest plot that shows the effect size estimates between various ethnic groups. We can see that generally, there is a lot of overlap in the effect sizes and their standard error ranges.

```{r Forest Plot Nonsig SNPs, echo = FALSE}
row_n <- 1

df_beta <- t(non_sig_gene[row_n, 5:9])
row.names(df_beta) <- c("AA", "EA", "EAS", "HA", "SA")
df_se <- t(non_sig_gene[row_n, 10:14])

chr_id <- non_sig_gene[row_n, 1]
pos_id <- non_sig_gene[row_n, 2]
snp_id <- paste("CHR: ", chr_id, ", ", "POS:", pos_id, sep = "")

mamba_val <- non_sig_gene[row_n, 3]
prp_val <- non_sig_gene[row_n, 4]
rep_val <- paste("MAMBA PPR: ", round(mamba_val, digits = 3), ", ", "PRP: ", round(prp_val, digits = 3), sep = "")

ggplot(as.data.frame(df_beta), aes(x = df_beta, y = row.names(df_beta))) +
  geom_pointrange(aes(xmin = df_beta - df_se, xmax = df_beta + df_se)) +
  ggtitle(snp_id, subtitle = rep_val) +
  xlab("Beta") +
  ylab("Ethnicity")
```


In contrast, below we have a SNP that the replicability packages have determined as not replicable. There is significantly less overlap between the estimated effect sizes and their standard error ranges for each ethnicity group.

```{r Forest Plot Significant SNPs, echo = FALSE}
row_n <- 1

df_beta <- t(sig_gene[row_n, 5:9])
row.names(df_beta) <- c("AA", "EA", "EAS", "HA", "SA")
df_se <- t(sig_gene[row_n, 10:14])

# SNP ID
chr_id <- sig_gene[row_n, 1]
pos_id <- sig_gene[row_n, 2]
snp_id <- paste("CHR: ", chr_id, ", ", "POS: ", pos_id, sep = "")

# PPR and PRP values
mamba_val <- sig_gene[row_n, 3]
prp_val <- sig_gene[row_n, 4]
rep_val <- paste("MAMBA PPR: ", round(mamba_val, digits = 3), ", ", "PRP: ", round(prp_val, digits = 3), sep = "")

ggplot(as.data.frame(df_beta), aes(x = df_beta, y = row.names(df_beta))) +
  geom_pointrange(aes(xmin = df_beta - df_se, xmax = df_beta + df_se)) +
  ggtitle(snp_id, subtitle = rep_val) +
  xlab("Beta") +
  ylab("Ethnicity")
```


A shiny app version of these plots is also provided that let's you compare different SNPs.


```{r, include = FALSE}
prp_mean <- mean(post_prp_data_pval)
prp_se <- sd(post_prp_data_pval)/sqrt(length(post_prp_data_pval))
mamba_mean <- mean(mamba_data$ppr)
mamba_se <- sd(mamba_data$ppr)/sqrt(length(mamba_data$ppr))


```

