---
title: "R Notebook, Week of 8/18"
output:
  html_document:
    df_print: paged
    toc: true
  pdf_document:
    toc: true
  html_notebook: default
editor_options:
  chunk_output_type: console
runtime: shiny
---

```{r, include=FALSE}
knitr::opts_chunk$set(
  warning = FALSE
)
```

```{r Libraries, include = FALSE}
# libraries and other things
library(PRP)
library(rstudioapi)
library(mamba)
library(data.table)
library(parallel)
library(knitr)
library(pander)
library(ggplot2)
library(tidyverse)
library(dplyr)
library(shiny)

#setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

set.seed(2021)
```

This file is a supplement to the 8-25 .Rmd notebook. It provides forest plots for various SNPs.

```{r Replicability Function, include = FALSE}
source(file = "data/giant_replicable.R")
source(file = "data/Inv_Var_Weight.R")
```

```{r Reading Data Function, include = FALSE}
get_val <- function(file_name, old_col, new_col) {
  # read in files and get values
  table_select <- read.table(file = file_name, header = TRUE) %>%
    select(c("CHR", "POS", old_col)) 
  
  for (i in 1:length(old_col)) {
    table_select <- table_select %>%
      rename(!!new_col[i] := !!old_col[i])
  }
  
  return(table_select)
}
```


# Giant Consortium Height

```{r, include = FALSE}
# loading in cleaned data, or if it doesn't exist then cleaning it now
if (file.exists(file = "data/giant_height/height_beta.rda") &&
    file.exists(file = "data/giant_height/height_se.rda")) {
  
  load(file = "data/giant_height/height_beta.rda")
  load(file = "data/giant_height/height_se.rda")
  
} else {
  
  # height datasets
  height_all <- get_val(file = "data/giant_height/height_All_add_SV.txt",
                        old_col = c("beta", "se", "Pvalue"),
                        new_col = c("beta_all", "se_all", "pval"))
  height_AA <- get_val(file = "data/giant_height/height_AA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_AA", "se_AA"))
  height_EA <- get_val(file = "data/giant_height/height_EA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_EA", "se_EA"))
  height_EAS <- get_val(file = "data/giant_height/height_EAS_add_SV.txt",
                        old_col = c("beta", "se"),
                        new_col = c("beta_EAS", "se_EAS"))
  height_HA <- get_val(file = "data/giant_height/height_HA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_HA", "se_HA"))
  height_SA <- get_val(file = "data/giant_height/height_SA_add_SV.txt",
                       old_col = c("beta", "se"),
                       new_col = c("beta_SA", "se_SA"))
  
  
  # put all values together
  height_list <- list(height_AA, height_EA, height_EAS, height_HA, height_SA, height_all)
  
  
  # merging data and dropping chr and pos columns
  height_vals <- Reduce(function(df1, df2) {
    merge(x = df1, y = df2, by = c("CHR", "POS"), all = TRUE)
  },
  height_list) %>%
    filter(pval <= 0.05/nrow(height_all))
  
  
  # remove any rows that are all NA
  height_vals[height_vals == Inf] <- NA
  
  
  # indices that are all NA values
  allnaind <- height_vals %>%
    select(-c("CHR", "POS", "beta_all", "se_all", "pval")) %>%
    apply(MARGIN=1, function(row_val) {
      all(is.na(row_val))
    } )
  
  height_vals <- height_vals[!allnaind, ] # exclude rows that are all NA
  
  
  # extracting beta and se and saving it
  beta_cols <- c("beta_AA", "beta_EA", "beta_EAS", "beta_HA", "beta_SA")
  height_beta <- height_vals %>%
    select("CHR", "POS", beta_cols)
  
  se_cols <- c("se_AA", "se_EA", "se_EAS", "se_HA", "se_SA")
  height_se <- height_vals %>%
    select("CHR", "POS", se_cols)
  
  
  save(height_beta, file = "data/giant_height/height_beta.rda")
  save(height_se, file = "data/giant_height/height_se.rda")  
  
}
```

```{r, include = FALSE}
if (file.exists(file = "data/giant_height/height_prp.rda") &&
    file.exists(file = "data/giant_height/height_prp_val.rda") &&
    file.exists(file = "data/giant_height/height_mamba.rda")) {
  
  load(file = "data/giant_height/height_prp.rda")
  load(file = "data/giant_height/height_prp_val.rda")
  load(file = "data/giant_height/height_mamba.rda")
  
} else {
  beta_cols <- c("beta_AA", "beta_EA", "beta_EAS", "beta_HA", "beta_SA")
  height_beta_val <- height_beta %>%
    select(beta_cols)
  
  se_cols <- c("se_AA", "se_EA", "se_EAS", "se_HA", "se_SA")
  height_se_val <- height_vals %>%
    select(se_cols)
  
  giant_replicable(beta = height_beta_val,
                   se = height_se_val,
                   prp_file = "data/giant_height/height_prp.rda",
                   prp_pval_file = "data/giant_height/height_prp_val.rda",
                   mamba_data_file = "data/giant_height/height_mamba.rda")
  
  load(file = "data/giant_height/height_prp.rda")
  load(file = "data/giant_height/height_prp_val.rda")
  load(file = "data/giant_height/height_mamba.rda")
  
}


```

```{r Replicability Data, include = FALSE}
# MAMBA
height_ppr_val <- as.data.frame(mamba_data$ppr) %>%
  rename("MAMBA" = "mamba_data$ppr")

# PRP
height_prp_pval <- as.data.frame(post_prp_data_pval) %>%
  rename("PRP" = "post_prp_data_pval")
```


```{r Height Gene Replicability Filter, include = FALSE}
# combined data
chrom_id <- height_beta %>%
  select("CHR", "POS")

# merge all data back together (IDs, )
height_rep_df <- cbind(chrom_id, height_ppr_val, height_prp_pval) %>%
  merge(y = height_beta, by = c("CHR", "POS"), all = TRUE) %>%
  merge(y = height_se, by = c("CHR", "POS"), all = TRUE)

# filter out values less than 0.05
# SNPs with MAMBA PPR and PRP values less than 0.05
sig_gene <- height_rep_df %>%
  filter(MAMBA <= 0.05, PRP <= 0.05)

# filter, keep nonsignificant SNPs
non_sig_gene <- height_rep_df %>%
  filter(MAMBA > 0.05 | PRP > 0.05)
```


We'll look at how the effect sizes look between SNPs that MAMBA and PRP have deemed significant and not signficant. Below, we have a forest plot that shows the effect size estimates between various ethnic groups. We can see that generally, there is a lot of overlap in the effect sizes and their standard error ranges.


```{r, echo = FALSE}
forest_plot_shiny <- function(snp, df) {
  df_beta <- t(df[snp, 5:9])
  row.names(df_beta) <- c("AA", "EA", "EAS", "HA", "SA")
  df_se <- t(df[snp, 10:14])
  
  mamba_val <- df[snp, 3]
  prp_val <- df[snp, 4]
  rep_val <- paste("MAMBA PPR: ", round(mamba_val, digits = 3), ", ", "PRP: ", round(prp_val, digits = 3), sep = "")
  
  ggplot(as.data.frame(df_beta), aes(x = df_beta, y = row.names(df_beta))) +
    geom_pointrange(aes(xmin = df_beta - df_se, xmax = df_beta + df_se)) +
    ggtitle(snp, subtitle = rep_val) +
    xlab("Beta") +
    ylab("Ethnicity")
}


# rename rows to SNPs
# first for nonsignificant SNPs
snp_list <- c()

for (i in 1:nrow(non_sig_gene)) {
  chr_id <- non_sig_gene[i, 1]
  pos_id <- non_sig_gene[i, 2]
  snp_id <- paste("CHR: ", chr_id, ", ", "POS: ", pos_id, sep = "")
  snp_list <- c(snp_list, snp_id)
  
}
row.names(non_sig_gene) <- snp_list


# next to significant SNPs
snp_list <- c()

for (i in 1:nrow(sig_gene)) {
  chr_id <- sig_gene[i, 1]
  pos_id <- sig_gene[i, 2]
  snp_id <- paste("CHR: ", chr_id, ", ", "POS: ", pos_id, sep = "")
  snp_list <- c(snp_list, snp_id)
  
}
row.names(sig_gene) <- snp_list
```

```{r Forest Plot Nonsignificant, echo = FALSE}
ui <- fluidPage(
  
  inputPanel(
    selectInput(inputId = "snp", 
                label = "SNP ID:",
                choices = rownames(non_sig_gene)
    )
   
  ),
   plotOutput("nonsig_snp_plot", height=500)
)

server <- function(input, output) {
  output$nonsig_snp_plot = renderPlot(forest_plot_shiny(input$snp, non_sig_gene))
}


shinyApp(ui, server, options = list(height = 625))

```

In contrast, below we have a SNP that the replicability packages have determined as not replicable. There is significantly less overlap between the estimated effect sizes and their standard error ranges for each ethnicity group.


```{r Forest Plot Significant, echo = FALSE}
ui <- fluidPage(
  
  inputPanel(
    selectInput(inputId = "snp", 
                label = "SNP ID:",
                choices = rownames(sig_gene)
    )
   
  ),
   plotOutput("sig_snp_plot", height=500)
)

server <- function(input, output) {
  output$sig_snp_plot = renderPlot(forest_plot_shiny(input$snp, sig_gene))
}


shinyApp(ui, server, options = list(height = 625))

```

