---
title: "R Notebook"
output: html_notebook
---

Using the data from the archive paper.

```{r}
library(devtools)
library(PRP)
library(mamba)
#> Loading required package: TMB
#> Loading required package: RcppEigen
library(data.table)
library(parallel)

set.seed(2021)

# change working directory to folder with current file
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```


```{r}
# loading data from mamba example file
load(file = "../test_data/ex_betajk.rda")
load(file = "../test_data/ex_sjk2.rda")

```

```{r}
mamba_beta_test <- data.frame(betajk[, c("b_1", "b_2")])
mamba_sjk_test <- data.frame(sjk2[, c("s2_1", "s2_2")])


beta_snp1 <- mamba_beta_test[, 1]
beta_snp2 <- mamba_beta_test[, 2]
beta_ex <- rbind(beta_snp1, beta_snp2)

sjk_snp1 <- mamba_sjk_test[, 1]
sjk_snp2 <- mamba_sjk_test[, 2]
sjk_ex <- rbind(sjk_snp1, sjk_snp2)
#beta_numeric <- as.numeric(unlist(mamba_beta_test), )
```

```{r}
posterior_prp(beta = mamba_beta_test[, 1],
              se = mamba_sjk_test[, 1])
```

Add a new chunk by clicking the *Insert Chunk* button on the toolbar or by pressing *Ctrl+Alt+I*.

When you save the notebook, an HTML file containing the code and output will be saved alongside it (click the *Preview* button or press *Ctrl+Shift+K* to preview the HTML file).

The preview shows you a rendered HTML copy of the contents of the editor. Consequently, unlike *Knit*, *Preview* does not run any R code chunks. Instead, the output of the chunk when it was last run in the editor is displayed.
